Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-01 ind_files/01_Introduction.tex
Streamline introduction

In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. Part \ref{part_theory} presents the mathematical formalism in a detailed, self-contained, pedagogical manner. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required. I present a brief introduction to the relevant mathematics in section \ref{sec_math_background} and present a toy example of tracking population size of a population of identical individuals in section \ref{sec_1D_processes}. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process, thus illustrating all the major tools required. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, use the system size expansion to arrive at a continuous description of change in trait frequencies as an SDE. Under mild assumptions, I show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations from stochastic first principles. While the mathematics of section \ref{sec_nD_processes} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here, though it has been used in several specific models of specific systems. As such, chapter \ref{sec_nD_processes} can be thought of as a conceptual synthesis and unification. Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck in terms of functional derivatives. Under mild assumptions, I show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues \citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, is entirely original both mathematically and biologically. Mathematically, the ideas present heuristic, accessible alternatives in terms of a `stochastic field theory' to the rigorous tools of martingale theory and measure-valued branching processes that are usually employed to describe the evolution of quantitative traits \citep{champagnat_unifying_2006,etheridge_mathematical_2011, week_white_2021}. This formulation in terms of stochastic field equations also opens up the model to analysis using the powerful heuristic tools of physics such as the path integral \citep{doi_second_1976, peliti_path_1985, dodd_many-body_2009, chow_path_2015, weber_master_2017}, though I do not attempt to leverage these connections here, leaving it as an opportunity for future study. Examples for one-dimensional, multi-dimensional, and quantitative trait models that can be analyzed in this formalism are presented in Appendix \ref{App_examples}.\\

>>>>

In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. Part \ref{part_theory} presents the mathematical formalism in a detailed, self-contained, pedagogical manner. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required. I present a brief introduction to the relevant mathematics in section \ref{sec_math_background} and present a toy example of tracking population size of a population of identical individuals in section \ref{sec_1D_processes}. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process, thus illustrating all the major tools required. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, use the system size expansion to arrive at a continuous description of change in trait frequencies as an SDE. Under mild assumptions, I show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations from stochastic first principles. While the mathematics of section \ref{sec_nD_processes} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects. As such, chapter \ref{sec_nD_processes} can be thought of as a tutorial, conceptual synthesis, and unification. Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck in terms of functional derivatives. Under mild assumptions, I show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues \citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, is entirely original both mathematically and biologically. Mathematically, the ideas present heuristic, accessible alternatives to the rigorous tools of martingale theory and measure-valued branching processes that are usually employed to describe the evolution of quantitative traits \citep{champagnat_unifying_2006,etheridge_mathematical_2011, week_white_2021} in terms of a `stochastic field theory'. This formulation in terms of stochastic field equations also opens up the model to analysis using the powerful heuristic tools of physics such as the path integral \citep{doi_second_1976, peliti_path_1985, dodd_many-body_2009, chow_path_2015, weber_master_2017}, though I do not attempt to leverage these connections here, leaving it as an opportunity for future study. Examples for one-dimensional, multi-dimensional, and quantitative trait models that can be analyzed in this formalism are presented in Appendix \ref{App_examples}.\\
